---
title: "Reproducible Research Project 1"
author: "Ramaranjan"
date: "Monday, June 08, 2015"
output: html_document
---

# Reproducible Research Project - 1

- ## Loading and preprocessing the data

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
setwd("C:/Users/rruj/Downloads")
data <- read.csv("activity.csv")
```

- ## What is mean total number of steps taken per day?

```{r, echo=TRUE, fig.height=4, fig.width=8, warning=FALSE, message=FALSE, fig.align='center'}
# Transforming the dataset to a desired format
raw <- tbl_df(data)
by_date <- group_by(raw, date)
steps_by_date <- summarize(by_date, sum(steps))
colnames(steps_by_date)[2] <- "steps"

#Plotting the graph using the ggplot2 package
g <- ggplot(steps_by_date, aes(x=steps))
g <- g + geom_histogram(binwidth=2000, color="red")
g <- g + labs(x="Sum of steps", title = "Distribution of steps")
g

mean_steps <- round(with(steps_by_date, mean(steps, na.rm=T)),2)
median_steps <- with(steps_by_date, median(steps, na.rm=T))
```

- *Mean Steps* : The mean steps taken each day is `r mean_steps`

- *Median Steps* : The median steps taken each day is `r median_steps`

```{r, warning=FALSE, message=FALSE, fig.height=4, fig.width=8, fig.align='center'}
#Plotting the time-series graph for 5-minute interval averaged across all days
by_interval <- group_by(raw, interval)
steps_by_interval <- summarize(by_interval, mean(steps, na.rm = TRUE))
colnames(steps_by_interval)[2] <- "steps"
g2 <- ggplot(steps_by_interval, aes(x=interval, y=steps))
g2 <- g2 + geom_line()
g2 <- g2 + labs(x="5-minute interval", y="Average step count", title="Time Distribution of step count")
g2

```

- ## Are there differences in activity patterns between weekdays and weekends?
```{r, warning=FALSE, message=FALSE, fig.height=4, fig.width=8, fig.align='center'}
# Adding a column for weektime
data$day <- with(data, weekdays(as.Date(date, format="%Y-%m-%d")))
data$weektime <- with(data, as.factor(ifelse(day %in% c("Sunday","Saturday"),"Weekend","Weekday")))
data <- data[,c(1:3,5)]

#Transforming the dataset
raw2 <- tbl_df(data)
by_weektime_interval <- group_by(raw2, weektime, interval)
avg_steps <- summarize(by_weektime_interval, mean(steps, na.rm = T))
colnames(avg_steps)[3] <- "Average.Steps"

# Plotting the activity patterns between weekends and weekdays
g3 <- ggplot(avg_steps, aes(x=interval, y=Average.Steps))
g3 <- g3 + geom_line()
g3 <- g3 + facet_wrap( ~ weektime)
g3

# Mean of steps in weekdays
mean(avg_steps$Average.Steps[avg_steps$weektime == "Weekday"], na.rm = T)

# Mean of steps in weekends
mean(avg_steps$Average.Steps[avg_steps$weektime == "Weekend"], na.rm = T)
```

### *Conslusion : It can be clearly observed from the graphs and the averages that more steps are taken during the weekends (43.07 mean steps) than the weekdays (35.33 mean steps)*