---
title: "Regression Models"
author: "Ramaranjan Ruj"
date: "June 16, 2015"
output: pdf_document
---

Motor Trend, a magazine for the automative industry is interested in exploring the relationship between a set of variables and miles per gallon (mpg) as the outcome. They are interested in the following questions, 
- "Is an automatic or manual transmission better for MPG"
- "Quantify the MPG difference between automatic and manual transmissions"

Given is a dataset named mtcars, the goal here is to develop a model with explains the relationship between the variables and mpg. We will build multiple regression models and select the one which has the best outcome.

```{r, warning=FALSE, message=FALSE, results='hide'}
# Loading the required libraries
library(ggplot2)
library(datasets)
library(gridExtra)

#Loading the dataset
data(mtcars)

#Converting the numeric variables to factors
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$vs <- as.factor(mtcars$vs)
mtcars$am <- as.factor(mtcars$am)
mtcars$gear <- as.factor(mtcars$gear)
mtcars$carb <- as.factor(mtcars$carb)
names(mtcars)[9] <- "Transmission"
levels(mtcars$Transmission) <- c("Automatic","Manual")
```

```{r}
#Regression Model-1 : mpg as outcome and 'am' as the variable
fit_am <- lm(mpg ~ Transmission, data=mtcars)
summary(fit_am)
```

- In this model, the transmission variable is significant as expected with a p-value of 0.000285. However, this model fails to explain the variance and explains only 33.8% of it. We now try to find a better model and include all variables.

```{r}
#Regression Model-2 : mpg as outcome with all input variables
fit_all <- lm(mpg ~ ., data=mtcars)
summary(fit_all)
```

- In this regression model, we include all variables. This model does explain the variance better (77.9%) but doesn't have any significant variable. We will try to change the model and find the significant variables with the backward selection technique.

```{r, results='hide'}
#Regression Model-3 : Backward selection to identify the significant variables
fit_back <- step(fit_all, k=log(nrow(mtcars)))
```

```{r}
summary(fit_back)
```

- With the backward technique, we have identified the variables wt, qsec and Transmission type as the significant variables. This model also explains most of the variance (83.3%). We will select this regression model for quantifying the mpg difference. 

### "Is an automatic or manual transmission better for MPG"
 - As per out third model, we expect a *improvement* in mpg with manual transmission. The boxplot also shows that the mean mpg for manual is better than the mean mpg for automatic transmission.
 
### "Quantify the MPG difference between automatic and manual transmissions"
 - While selecting the 3rd model, the difference in mpg for manual transmission and automatic transmission is 2.93 mpg ( Here the weight wt(lbs/1000) and qsec is kept constant).
 
### Figures:

```{r, fig.width=12}
# Plot for the relation between mpg and Transmission
g <- ggplot(mtcars, aes(x=Transmission, y=mpg))
g <- g + geom_boxplot(aes(fill=Transmission))
g <- g + labs(title = "Boxplot for mpg v/s Transmission")

g2 <- ggplot(mtcars, aes(x=wt, y=mpg))
g2 <- g2 + geom_point(aes(colour=cyl))
g2 <- g2 + geom_smooth(method=lm, aes(colour=cyl))
g2 <- g2 + labs(title="Plot of mpg v/s wt")

grid.arrange(g, g2, ncol=2)
```

```{r, fig.width=12, fig.height=8}
# Plot for for final selection model
par(mfrow = c(2,2))
plot(fit_back)
```